---
title: "Global BF-NEE trends"
---

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(scales)

dir_diag <- "~/Symobio/Data/Processed/BF_NEE_Diagnostics"

diag <- read_csv(
  file.path(
    dir_diag,
    "bfnee_country_diagnostics_all_transitions.csv"
  ),
  show_col_types = FALSE
)

valid <- read_csv(
  file.path(
    dir_diag,
    "bfnee_valid_countries.csv"
  ),
  show_col_types = FALSE
)

valid_iso3 <- valid %>%
  filter(valid_for_analysis) %>%
  pull(iso3)
```

# Overview

This page summarizes global BF-NEE trends based on countries that passed all processing and scale diagnostics.

```{r}
global_trend <- diag %>%
  filter(
    iso3 %in% valid_iso3,
    status == "ok",
    suspicious_old_scale == FALSE
  ) %>%
  group_by(year, transition) %>%
  summarise(
    global_bfnee_km2 =
      sum(bfnee_sum_km2, na.rm = TRUE),

    global_natural_loss_km2 =
      sum(natural_loss_sum_km2, na.rm = TRUE),

    n_countries =
      n_distinct(iso3),

    .groups = "drop"
  ) %>%
  arrange(year) %>%
  mutate(
    bfnee_index =
      global_bfnee_km2 /
      first(global_bfnee_km2) * 100,

    natural_loss_index =
      global_natural_loss_km2 /
      first(global_natural_loss_km2) * 100,

    bfnee_share_of_loss =
      global_bfnee_km2 /
      global_natural_loss_km2
  )
```

# Global BF-NEE

```{r}
ggplot(
  global_trend,
  aes(
    x = year,
    y = global_bfnee_km2
  )
) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    title = "Global BF-NEE",
    subtitle = "Validated countries only",
    x = "Year",
    y = "BF-NEE (km²)"
  ) +
  theme_minimal()
```

# Global natural ecosystem loss

```{r}
ggplot(
  global_trend,
  aes(
    x = year,
    y = global_natural_loss_km2
  )
) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    title = "Global natural ecosystem loss",
    subtitle = "Validated countries only",
    x = "Year",
    y = "Natural ecosystem loss (km²)"
  ) +
  theme_minimal()
```

# Absolute comparison

```{r}
global_trend_long <- global_trend %>%
  select(
    year,
    global_bfnee_km2,
    global_natural_loss_km2
  ) %>%
  pivot_longer(
    cols = c(
      global_bfnee_km2,
      global_natural_loss_km2
    ),
    names_to = "indicator",
    values_to = "value_km2"
  ) %>%
  mutate(
    indicator = recode(
      indicator,
      global_bfnee_km2 =
        "BF-NEE",
      global_natural_loss_km2 =
        "Natural ecosystem loss"
    )
  )
```

```{r}
ggplot(
  global_trend_long,
  aes(
    x = year,
    y = value_km2,
    linetype = indicator
  )
) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    title = "Absolute comparison",
    subtitle = "Validated countries only",
    x = "Year",
    y = "Area (km²)",
    linetype = "Indicator"
  ) +
  theme_minimal()
```

# Relative change (first period = 100)

This figure compares rates of change only.

Both indicators are normalized so that the first period equals 100.

```{r}
global_trend_index_long <- global_trend %>%
  select(
    year,
    bfnee_index,
    natural_loss_index
  ) %>%
  pivot_longer(
    cols = c(
      bfnee_index,
      natural_loss_index
    ),
    names_to = "indicator",
    values_to = "index"
  ) %>%
  mutate(
    indicator = recode(
      indicator,
      bfnee_index =
        "BF-NEE",
      natural_loss_index =
        "Natural ecosystem loss"
    )
  )
```

```{r}
ggplot(
  global_trend_index_long,
  aes(
    x = year,
    y = index,
    linetype = indicator
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Relative change in BF-NEE and natural ecosystem loss",
    subtitle = "First period = 100",
    x = "Year",
    y = "Index",
    linetype = "Indicator"
  ) +
  theme_minimal()
```

# BF-NEE share of natural ecosystem loss

This figure shows the proportion of natural ecosystem loss represented by BF-NEE.

```{r}
ggplot(
  global_trend,
  aes(
    x = year,
    y = bfnee_share_of_loss
  )
) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    labels = percent_format(
      accuracy = 1
    )
  ) +
  labs(
    title = "BF-NEE as share of natural ecosystem loss",
    subtitle = "Validated countries only",
    x = "Year",
    y = "BF-NEE / natural ecosystem loss"
  ) +
  theme_minimal()
```

# Diagnostic table

```{r}
global_trend
```