---
title: "Regional regressions"
---

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(broom)
library(DT)
library(scales)

df <- read_csv(
  "data/bfnee_regression_dataset.csv",
  show_col_types = FALSE
)

df_model <- df %>%
  filter(
    !is.na(continent),
    !is.na(log_bfnee),
    !is.na(log_natural_loss),
    !is.na(log_weight_sum),
    natural_loss_sum_km2 > 0,
    bfnee_sum_km2 > 0
  )
```

# Data coverage

```{r}
df_model %>%
  distinct(continent, iso3) %>%
  count(continent, name = "n_countries") %>%
  arrange(continent)
```

```{r}
df_model %>%
  count(continent, year) %>%
  arrange(continent, year)
```

# Model specification

The first regional consistency model is:

```text
log(BF-NEE + 1) ~ log(natural ecosystem loss + 1) + log(weight sum + 1)
```

where `weight_sum` represents the aggregated commodity-consumption and production-attribution component.

# BF-NEE and natural ecosystem loss

```{r}
ggplot(
  df_model,
  aes(
    x = natural_loss_sum_km2,
    y = bfnee_sum_km2
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ continent, scales = "free") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "BF-NEE vs natural ecosystem loss",
    subtitle = "Validated countries only",
    x = "Natural ecosystem loss (km²)",
    y = "BF-NEE (km²)"
  ) +
  theme_minimal()
```

# Log-log relationship

```{r}
ggplot(
  df_model,
  aes(
    x = log_natural_loss,
    y = log_bfnee
  )
) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ continent, scales = "free") +
  labs(
    title = "Log-log relationship between BF-NEE and natural ecosystem loss",
    subtitle = "Validated countries only",
    x = "log(natural ecosystem loss + 1)",
    y = "log(BF-NEE + 1)"
  ) +
  theme_minimal()
```

# Regional regression models

```{r}
fit_region <- function(x) {
  lm(
    log_bfnee ~ log_natural_loss + log_weight_sum,
    data = x
  )
}

regional_models <- df_model %>%
  group_by(continent) %>%
  group_modify(~ {
    model <- fit_region(.x)
    
    tidy_model <- tidy(model)
    glance_model <- glance(model)
    
    tidy_model %>%
      mutate(
        r_squared = glance_model$r.squared,
        adj_r_squared = glance_model$adj.r.squared,
        n_obs = glance_model$nobs
      )
  }) %>%
  ungroup()

regional_models
```

```{r}
datatable(
  regional_models %>%
    arrange(continent, term),
  options = list(pageLength = 25)
)
```

# Coefficient comparison

```{r}
regional_coefficients <- regional_models %>%
  filter(term != "(Intercept)")

ggplot(
  regional_coefficients,
  aes(
    x = continent,
    y = estimate
  )
) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = estimate - 1.96 * std.error,
      ymax = estimate + 1.96 * std.error
    ),
    width = 0.15
  ) +
  facet_wrap(~ term, scales = "free_y") +
  labs(
    title = "Regional regression coefficients",
    subtitle = "Error bars show approximate 95% confidence intervals",
    x = "Region",
    y = "Coefficient estimate"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

# Model fit by region

```{r}
model_fit <- regional_models %>%
  distinct(
    continent,
    r_squared,
    adj_r_squared,
    n_obs
  )

ggplot(
  model_fit,
  aes(
    x = reorder(continent, r_squared),
    y = r_squared
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Model fit by region",
    x = "Region",
    y = "R²"
  ) +
  theme_minimal()
```

# Residual diagnostics

```{r}
regional_augmented <- df_model %>%
  group_by(continent) %>%
  group_modify(~ {
    model <- fit_region(.x)
    augment(model, data = .x)
  }) %>%
  ungroup()
```

```{r}
ggplot(
  regional_augmented,
  aes(
    x = .fitted,
    y = .resid
  )
) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ continent, scales = "free") +
  labs(
    title = "Residual diagnostics by region",
    x = "Fitted log(BF-NEE + 1)",
    y = "Residual"
  ) +
  theme_minimal()
```

# Largest positive residuals

Countries with large positive residuals have higher BF-NEE than expected from natural ecosystem loss and the aggregate attribution weight.

```{r}
regional_augmented %>%
  arrange(desc(.resid)) %>%
  select(
    continent,
    iso3,
    country_name,
    transition,
    year,
    bfnee_sum_km2,
    natural_loss_sum_km2,
    weight_sum,
    .fitted,
    .resid
  ) %>%
  head(30) %>%
  datatable(
    options = list(pageLength = 15)
  )
```

# Largest negative residuals

Countries with large negative residuals have lower BF-NEE than expected from natural ecosystem loss and the aggregate attribution weight.

```{r}
regional_augmented %>%
  arrange(.resid) %>%
  select(
    continent,
    iso3,
    country_name,
    transition,
    year,
    bfnee_sum_km2,
    natural_loss_sum_km2,
    weight_sum,
    .fitted,
    .resid
  ) %>%
  head(30) %>%
  datatable(
    options = list(pageLength = 15)
  )
```